Query Optimization for Dynamic Imputation

### 2017-05-02 rebuttal.tex

\item - S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data. I am slightly concerned that by focusing entirely on size, this might lead to skew in the training set. For example, consider a one-to-many join \verb/µ(R) |X| S/. Your optimization algorithm would likely wish to move imputation up: \verb/µ(R |X| S)/. Yes, there is more training data, but if the fanout for each tuple is skewed, some tuples in R will become over-represented in the join result. A similar observation holds for selection. For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree $\geq$ 0" presumably indicate invalid data that should not be used for training. (is this perhaps why you have so much error for full imputation on Q8?)

  % \resp{This is a reasonable concern. Optimizing for the maximum training set is flawed when the training data is skewed. We think that this problem is avoidable with a minor tweak to the loss function. Instead of optimizing for a large training set, we can optimize for a \emph{diverse} training set by tracking the

\item - S3.4.4: It appears that Time() assumes that plans are IO-bound. This is a perfectly reasonable assumption, but should be stated explicitly, as datasets that fit in memory will require a different cost model.

  \resp{Agreed. Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function. We have clarified this in the paper.}

@@

\item - S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator. Why not just take the same approach as selection and treat it as having negligible cost?

  \resp{This was an error in the paper. In the implementation, we do treat Drop as taking negligible time.}

### 2017-05-02 paper/algorithm.tex
paper: make figure smaller
Add a couple points to rebuttal
clean up algorithms

\begin{figure}
\Tree [.$\pi$ [.$g$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
\caption{Query plan schematic for the type of traditional plans explored (absent imputation operators).}
\label{fig:query-schematic}
\end{figure}

@@

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Loss}(q), \textsc{Time}(q) \rangle$.
$\textsc{Loss}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.

@@

This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Loss}$ for the current query.

In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Loss}$ value that the user is willing
to tolerate relative to the minimum possible amongst frontier plans.

\begin{definition}
Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
We say a plan $p^\alpha$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p')$. 
\end{definition}

@@

\subsubsection{Cardinality estimation}
The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ rely on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.

@@

\subsubsection{Imputation quality}
We use these histogram estimates to compute $\text{Loss}(Q)$, a measure averaged over imputation operations, as follows.

\begin{align*}
  \textsc{L}(q) &= \begin{cases}
     1 + \textsc{L}(q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}} + \textsc{L}(q')  & q = \mu_C(q') \\
    \textsc{L}(q_l') + \textsc{L}(q_r') & q = q_l' \bowtie_\psi q_r' \\
    \textsc{L}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = \text{some table } t\\
  \end{cases}\\
    \textsc{N}(q) &= \begin{cases}
     1 + \textsc{N}(q') & q = \delta_C(q') \\
     1 + \textsc{N}(q')  & q = \mu_C(q') \\
    \textsc{N}(q_l') + \textsc{N}(q_r') & q = q_l' \bowtie_\psi q_r' \\
    \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = \text{some table } t\\
  \end{cases}\\
  \textsc{Loss}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
\end{align*}

Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties, and $N$ counts the number of imputation operations.
For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
the product of the number of attributes and the number of tuples. Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. We encode the fact that the number of training observations has diminishing returns for imputation quality,
and find that scaling the cell count by applying square root works well in practice.

@@

Scan costs are estimated based on the number of tuples, 
page size, and a unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations.

We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page. The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the implementation of our system,
we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Loss}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Loss}, \textsc{Time} \rangle$
as a total order, allowing us to appropriately select the frontier. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier produced by the optimizer corresponds to the collection of plans associated with the empty dirty set.

@@

\begin{algorithm}
\input{algorithms/plan-helpers.tex}
\end{algorithm}

@@

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and loss functions (\Cref{sec:cost-model}) which describe the new algorithm.
For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
attributes $C$ in a relation $R$, returning a new relation $R'$.
The flexibility of this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation.

>>>>

\begin{figure}
  \begin{minipage}[c]{0.5\columnwidth}
    \Tree [.$\pi$ [.$g$ [.$\bowtie$ [.$\bowtie$ [.$\sigma$ $t_1$ ] [.$\sigma$ $t_2$ ] ] [.$\sigma$ $t_3$ ] ] ] ]
  \end{minipage}\begin{minipage}[c]{0.5\columnwidth}
    \caption{
       Query plan schematic for the type of traditional plans explored (absent imputation operators).
    } \label{fig:query-schematic}
  \end{minipage}
\end{figure}

@@

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is expressed as a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
$\textsc{Penalty}(q) \in [0,1]$ is an estimate of the amount of information lost by the imputation procedures used.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics and selectivity estimation of the query predicates.

@@

This set $P$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.

In order to pick a final plan from the frontier, our model introduces a trade-off parameter $\alpha$, which is an upper bound on the $\textsc{Penalty}$ value that the user is willing
to tolerate relative to the minimum possible amongst frontier plans.

\begin{definition}
Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Penalty}(p) - \min_{p' \in P}(\textsc{Penalty}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
We say a plan $p^\alpha$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p')$. 
\end{definition}

@@

\subsubsection{Cardinality estimation}
The computation of $\textsc{Time}(q)$ and $\textsc{Penalty}(q)$ rely on having an accurate cardinality estimate of a query plan $q$ and its sub-plans.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.

@@

\subsubsection{Imputation quality}
We estimate the quality of the imputations in a plan $Q$ using a heuristic measure $\text{Penalty}(Q)$, which is computed as follows.

\begin{align*}
  \textsc{L}(q) &= \begin{cases}
     1 + \textsc{L}(q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{ \textsc{Attr}(q') * \textsc{Card}(q')}} + \textsc{L}(q')  & q = \mu_C(q') \\
    \textsc{L}(q_l') + \textsc{L}(q_r') & q = q_l' \bowtie_\psi q_r' \\
    \textsc{L}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = \text{some table } t\\
  \end{cases}\\
    \textsc{N}(q) &= \begin{cases}
     1 + \textsc{N}(q') & q = \delta_C(q') \\
     1 + \textsc{N}(q')  & q = \mu_C(q') \\
    \textsc{N}(q_l') + \textsc{N}(q_r') & q = q_l' \bowtie_\psi q_r' \\
    \textsc{N}(q') & q = \sigma_\phi(q'), q = \pi_C(q') \\
    0 & q = \text{some table } t\\
  \end{cases}\\
  \textsc{Penalty}(q) &= \frac{\textsc{L}(q)}{\textsc{N}(q)}
\end{align*}

Each imputation operator incurs an individual penalty.
We define \textsc{Penalty} as the average over the penalties incurred by the imputation operators. Our recursive definition of $L$ accumulates these penalties and $N$ counts the number of imputation operations.
For example, $\delta$ contributes a penalty of 1 (i.e. we lose all information  that dropped tuples might have encoded).
Imputing \nullv{} fields with $\mu$ incurs a penalty $p \in (0, 1]$.
In this case, the penalty is inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation accuracy \textit{increases} when more complete attributes and complete tuples are available; correspondingly, the penalty \textit{decreases} with more values to work with.
Scaling the cell count by applying square root incorporates the intuition that the number of training observations has diminishing returns for imputation quality and we find that it works well in practice.

@@

Scan costs are estimated based on the number of tuples, 
page size, and a unit-less I/O cost per page. 
We assume that I/O costs dominate CPU costs for relational operators. 
A database implementation for which this assumption is not true, i.e. an in memory database, will need a different cost model.
Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations.

We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a filter, so its time complexity is correspondingly negligible. The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the implementation of our system,
we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Penalty}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Penalty}, \textsc{Time} \rangle$
as a total order, allowing us to appropriately select the frontier. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier produced by the optimizer corresponds to the collection of plans associated with the empty dirty set.

@@

@@

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and penalty functions (\Cref{sec:cost-model}) which describe the new algorithm.
For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
attributes $C$ in a relation $R$, returning a new relation $R'$.
The flexibility of this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation.

### 2017-05-02 paper/algorithms/plan-toplevel.tex
clean up algorithms

% \begin{algorithm}
\begin{algorithmic}[1]
  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information lost}
  \Ensure{An optimized query plan.}

  \Function{Plan}{$T, F, J, P, G, A$}
  \State $C_g \gets \bigcup_{\phi \in F} \Call{Attr}{\phi}$ \Comment{Collect relevant attributes.}~\label{lst:line:attr-start}
  \State $C_g \gets \bigcup_{\psi \in J} \Call{Attr}{\psi} \cup P \cup G \cup \Call{Attr}{A} \cup C_g$ \label{lst:line:attr-end}
  \State Let $Q$ be an empty plan cache.
  \For{$t \in T$} \Comment{Add selections to the plan cache.}
  \If {$\exists \phi: (t, \phi) \in F$}
  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, C_g}$~\label{lst:line:sel}
  \Else
  \State $Q[\{t\}] \lhd \{t\}$~\label{lst:line:scan}
  \EndIf
  \EndFor

  \State $\Call{OptJoin}{Q, T, J, C_g}$	\Comment{Optimize joins.}~\label{lst:line:join}
  \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}~\label{lst:line:group}
  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_g}$
  \EndIf

  \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}
  \State \Return $p \in B$ s.t. $p$ is $\alpha$-bound optimal.
  \EndFunction

\end{algorithmic}
\caption{Top-level query planner with imputations.}
\label{algo:top-level-planner}
% \end{algorithm}

>>>>

\newcommand{\InlineIte}[3]{#2\ \textbf{if}\ #1\ \textbf{else}\ #3}
\renewcommand{\algorithmicindent}{1em}

\begin{algorithmic}[1]
    \Require{
    \Statex
    \begin{itemize}[leftmargin=*,noitemsep]
    \item $q$: A query plan.
    \item $C_l$: A set of attributes that must be imputed in the output of this query plan.
    \item $C_g$: The set of attributes which are used in the final plan.
    \end{itemize}}
  \Function{Impute}{$q,\ C_l,\ C_g$}
  \State $D_{must} \gets \Call{Dirty}{q} \cap C_l$
  \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_g)$
  \State $Q \gets (\InlineIte{D_{must} = \emptyset}{\{q\}}{\{\mu_{D_{must}}(q),\ \delta_{D_{must}}(q)\}})$
  \State \Return $(\InlineIte{D_{may} = \emptyset}{Q}{Q \cup \{\mu_{D_{may}}(q), \delta_{D_{may}}(q) \}})$
  \EndFunction

  \Statex

  \Require{
    \Statex
    \begin{itemize}[leftmargin=*,noitemsep]
    \item $T$: A set of tables.
    \item $F$: A $T \times \Phi$ relation between tables and filter predicates.
    \item $J$: A $T \times \Psi \times T$ relation between tables and join predicates.
    \item $P$: A set of projection attributes.
    \item $G$: A set of grouping attributes.
    \item $A$: An aggregation function.
    \item $\alpha$: A parameter in $[0, 1]$ that expresses the trade-off between performance and imputation quality.
    \end{itemize}}
  \Function{Plan}{$T,\ F,\ J,\ P,\ G,\ A,\ \alpha$}
  \State $C_g \gets \bigcup_{\psi \in J} \Call{Attr}{\psi}$ \Comment{Collect relevant attributes.}~\label{lst:line:attr-start} 
  \State $C_g \gets C_g\ \cup\ \bigcup_{\phi \in F} \Call{Attr}{\phi}\ \cup\ P \cup\ G\ \cup\ \Call{Attr}{A}$~\label{lst:line:attr-end}
  \Statex
  \State Let $Q$ be an empty plan cache.
  \For{$t \in T$} \Comment{Add selections to the plan cache.}
  \If {$\exists \phi: (t, \phi) \in F$}
  \State $Q[\{t\}] \lhd \{\sigma_{\phi}(q)\ |\ q \in \Call{Impute}{t, \Call{Attrs}{\phi}, C_g}\}$~\label{lst:line:sel}
  \Else\ $Q[\{t\}] \lhd \{t\}$~\label{lst:line:scan}
  \EndIf
  \EndFor
  \Statex
  \For{$size \in 2\dots|T|$} \Comment{Optimize joins.}~\label{lst:line:join}
  \For{$S \in \{\text{all length}\ size\ \text{subsets of}\ T\}$}
  \For{$t \in S$}
  \For{$(t,\ \psi,\ t') \in J$ where $t' \in S \setminus t$}
  \State $L \gets \{\Call{Impute}{q,\ \Call{Attrs}{\psi},\ C_g} ~|~ q \in Q[S \setminus t] \}$
  \State $R \gets \{\Call{Impute}{q,\ \Call{Attrs}{\psi},\ C_g} ~|~ q \in Q[\{t\}] \}$
  \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L,\ r \in R\}$
  \EndFor
  \EndFor
  \EndFor
  \EndFor
  \Statex
  \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}~\label{lst:line:group}
  \State $C_l \gets G \cup \Call{Attrs}{A}$
  \State $B \gets \bigcup_{q \in B} \{g(q', G, A) ~|~ q' \in \Call{Impute}{q, C_l,\ C_g}\}$
  \EndIf

  \State $B \gets \bigcup_{q \in B} \{\pi_P(q') ~|~ q' \in \Call{Impute}{q,\ P,\ P}\}$ \Comment{Add projections.}
  \State \Return $p \in B$ s.t. $p$ is $\alpha$-bound optimal.
  \EndFunction

\end{algorithmic}
\caption{A query planner with imputations.}
\label{algo:top-level-planner}
